High-Velocity Outflows Without AGN Feedback: \\ Eddington-Limited Star Formation in Compact Massive Galaxies

### 2015-07-17 doc/tex/sigmasfr/ms.tex
moved the tex and proposals directories to doc

\documentclass[apj]{emulateapj}

@@

\altaffiltext{1}{Center for Astrophysics and Space Sciences,
  University of California, San Diego, La Jolla, CA 92093, USA}
\altaffiltext{2}{Center for Galaxy Evolution Fellow}
\altaffiltext{3}{Department of Astronomy, University of
  Wisconsin-Madison, Madison, WI 53706, USA}
\altaffiltext{4}{Department of Physics and Astronomy, Dartmouth
  College, Hanover, NH 03755, USA}

@@

However, it also plausible that feedback from a compact starburst
could expel gas with such large velocities.  Indeed, there is evidence
for a positive correlation between outflow velocity and starburst
luminosity \citep{mar05,rup05,tre07}, albeit with significant scatter.
Furthermore, \citet{hec11} recently found outflows with maximum
velocities reaching $1500$~\kms\ in a sample of local starbursts with
compact nuclei, and argued that such velocities could be explained by
a wind that is launched from $r_0\sim100$~pc and driven by feedback
from massive stars and supernovae.

@@

% we focus our analysis on the galaxies with the most extreme
% $\sigmasfr$ values.

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{f1.ps}
\caption{Left: Observed UV--IR SEDs ($\lrest=0.1$--15~$\mu$m) for
  three galaxies with the largest SFR surface densities
  ($\sigmasfr\approx3000$~\units).  The top two SEDs are offset by 5
  and 10 magnitudes, respectively.  We show stellar population fits to
  the $\lrest=0.1$--3~$\mu$m emission (black solid line) and three
  templates for dust emission \citep[M82 starburst, blue dashed line;
    Arp 220 starburst, green dot-dashed line; obscured quasar, red
    dotted line;][]{pol07} scaled to match the $\lobs=4.6~\mu$m band.
  The starburst templates provide reasonable fits to the $\lobs=12$
  and 22~$\mu$m WISE photometry, while the quasar template does not.
  Right: HST/WFC3 F814W images (probing $\lrest\approx5000$~\AA)
  showing that these galaxies are dominated by a compact nucleus.}
\label{fig:seds}
\end{figure}

@@

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{f2.ps}
\caption{One-dimensional surface brightness profile for J0905+5759,
  which has the smallest effective radius in the sample.  The observed
  profile is shown as the solid black line, and the profiles of six
  stars in the same image are shown in blue, normalized to the same
  central surface brightness.  The best-fit de Vaucouleurs profile
  with $r_e=100$~pc is shown as a dashed red line, and for comparison
  a broader profile with $r_e=0.04\arcsec=290$~pc is shown as the
  dotted green line.  This galaxy is quite compact, but is more
  extended than a point source.}
\label{fig:profile}
\end{figure}

@@

% We gathered photometry from the All-Sky Release of the Wide-field
% Infrared Survey Explorer \citep[WISE,][]{wri10}, the Seventh Data
% Release of the Sloan Digital Sky Survey \citep[SDSS,][]{aba09}, and
% General Release 6 from the Galaxy Evolution Explorer
% \citep[GALEX,][]{mar05galex}.  
We gathered photometry from the WISE All-Sky Release, the SDSS Seventh
Data Release, and GALEX General Release 6.  We also obtained
$5\times30$~sec dithered exposures at 3.6~$\mu$m and 4.5~$\mu$m for
all sources with the Infrared Array Camera on the Spitzer Space
Telescope as part of GO program 60145.  We used the post--basic
calibrated data to perform aperture photometry on all sources and
point-source photometry on sources in crowded fields.  We show
spectral energy distributions (SEDs) for the three highest $\sigmasfr$
galaxies in Figure~\ref{fig:seds}.

% All the photometry was corrected for Galactic extinction
% based on the \citet{sch98} dust maps.

@@

\begin{figure}[!t]
\includegraphics[angle=0,scale=0.41]{f3.ps}
\caption{Spectra covering $\lrest=2500$--5200~\AA\ for the three
  galaxies shown in Figure~\ref{fig:seds}.  For clarity the top two
  spectra are offset by $+10$ and $+20$ units.  These spectra are
  dominated by the light of a young stellar population but have
  relatively weak nebular emission lines (\oii~$\lambda3727$,
  H$\beta$~$\lambda4861$, \oiii~$\lambda5007$) and strong
  \mgii~$\lambda\lambda2796,2803$ absorption arising from the
  interstellar medium.  The absense of broad \mgii\ or H$\beta$
  emission lines rules out a significant contribution to the optical
  continuum light from an AGN.  The panels on the right highlight the
  region around the \mgii\ doublet in both wavelength and velocity
  space to illustrate the outflow kinematics.  The spectrum in the
  bottom panel has sufficient spectral resolution
  ($\textnormal{FWHM}\approx8$~\kms) to resolve the intrinsic shape of
  the absorption-line profile, revealing that the gas near the
  centroid velocity ($v=-2470$~\kms) covers the entire galaxy.}
\label{fig:spectra}
\end{figure}

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.8]{f4.ps} 
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, with symbol size
  proportional to outflow velocity), along with samples of $z<0.3$
  gas-rich mergers (orange squares) and $z\sim1$ star-forming galaxies
  (shown with 68\%, 95\%, and 99.7\% contours).  We mark the empirical
  threshold for launching winds \citep[dotted line,
    $\sigmasfr\approx0.1$~\units;][]{hec02}, the 90th-percentile
  starburst intensity limit from \citet{meu97} (dashed line,
  $\sigmasfr\approx25$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx3000$~\units;][]{mur05,tho05,hop10}.  The
  representative error bar in the top-right portion of the plot
  corresponds to uncertainties of 0.3~dex in $\sigmasfr$ and 0.2~dex
  in stellar mass.  Our HST--WISE sample overlaps with the region
  characterized by gas-rich mergers, and extends to very large SFR
  surface densities near the Eddington limit, suggesting growth that
  is limited by momentum injection from massive stars.}
\label{fig:sigmasfr}
\end{center}
\end{figure*}

@@

While the SEDs (Figure~\ref{fig:seds}) and optical spectra
(Figure~\ref{fig:spectra}) for our sample indicate that their
bolometric output is dominated by star formation, it is worthwhile to
consider the level of ongoing AGN activity and its effect on our
measurements.  Among the high-$\sigmasfr$ galaxies, the strongest case
for detectable AGN emission can be made for J1506+5402, which has the
most luminous \oiii~$\lambda5007$ line in the HST sample.  This source
also has a weak \nev~$\lambda3426$ emission line, which is associated
with AGN activity \citep{gil10}.  It was observed with the Chandra
X-ray Observatory (proposal ID 11700896) and had four detected counts,
corresponding to a 0.5--8.0~keV X-ray luminosity of
$10^{42.7}$~erg~s$^{-1}$ (see P. Sell et al., in preparation for more
details on the 12/29 galaxies with Chandra observations).  Using the
relationship between 2--10~keV X-ray and 12.3~$\mu$m mid-IR luminosity
for AGNs from \citet{gan09}, one would expect a source with
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{42}$~erg~s$^{-1}$ to
have
$L_{\textnormal{\scriptsize{MIR}}}\approx7\times10^{42}$~erg~s$^{-1}$,
which is 400$\times$ smaller than observed luminosity for J1506+5402
($L_{\textnormal{\scriptsize{MIR}}}\approx3\times10^{45}$~erg~s$^{-1}$).
Based on its \oiii\ luminosity ($10^{42.1}$~erg~s$^{-1}$, which may
also include a contribution from star formation) and the calibration
for type 1 AGNs from \citet{hec05}, one would expect an intrinsic
2--10~keV luminosity of
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{43}$~erg~s$^{-1}$,
suggesting absorption by a factor of $\sim10$ in the X-rays.  However,
even if the X-ray attenuation were a factor of 100, which is the
typical value for local Compton-thick AGNs \citep{dia09}, the expected
mid-IR AGN contribution would only be relevant at the $\lesssim$30\%
level.  We conclude that the bolometric output of the galaxies in our
sample is dominated by star formation and that our results regarding
large $\sigmasfr$ values are not affected by AGN contamination.

% Thus, similar to local ULIRGs \citep[e.g.,][]{far07}, the
% bolometric output of the galaxies in our sample is dominated by star
% formation, and we conclude that our results regarding large
% $\sigmasfr$ values are not affected by AGN contamination.

@@

We acknowledge useful discussions with and assistance from James Aird,
Brandon Kelly, Dusan Keres, David Law, Alexander Mendez, Kate Rubin,
Art Wolfe, and Stijn Wuyts.  AMD acknowledges support from the
Southern California Center for Galaxy Evolution, a multi-campus
research program funded by the University of California Office of
Research.

@@

%\bibitem[Abazajian et al.(2009)]{aba09} Abazajian, K.~N.,
%  Adelman-McCarthy, J.~K., Ag{\"u}eros, M.~A., et al.\ 2009, \apjs,
%  182, 543
%\bibitem[Abel \& Satyapal(2008)]{abe08} Abel, N.~P., \& Satyapal,
%  S.\ 2008, \apj, 678, 686

@@

%\bibitem[Dale \& Helou(2002)]{dal02} Dale, D.~A., \& Helou, G.\ 2002,
%  \apj, 576, 159
%\bibitem[de Vaucouleurs(1948)]{dev48} de Vaucouleurs, G.\ 1948,
%  Annales d'Astrophysique, 11, 247
%\bibitem[Debuhr et al.(2012)]{deb12} Debuhr, J., Quataert, E., \& Ma,
%  C.-P.\ 2012, \mnras, 420, 2221

@@

%\bibitem[Farrah et al.(2007)]{far07} Farrah, D., Bernard-Salas, J.,
%  Spoon, H.~W.~W., et al.\ 2007, \apj, 667, 149
%\bibitem[Fazio et al.(2004)]{faz04} Fazio, G.~G., Hora, J.~L., Allen,
%x  L.~E., et al.\ 2004, \apjs, 154, 10

@@

%\bibitem[Heckman et al.(1993)]{hec93} Heckman, T.~M., Lehnert, M.~D.,
%  \& Armus, L.\ 1993, The Environment and Evolution of Galaxies, 188,
%  455

@@

%\bibitem[Martin et al.(2005)]{mar05galex} Martin, D.~C., Fanson, J.,
%  Schiminovich, D., et al.\ 2005, \apjl, 619, L1

@@

%\bibitem[Peng et al.(2010)]{pen10} Peng, C.~Y., Ho, L.~C., Impey,
%  C.~D., \& Rix, H.-W.\ 2010, \aj, 139, 2097

@@

%\bibitem[Schlegel et al.(1998)]{sch98} Schlegel, D.~J., Finkbeiner,
%  D.~P., \& Davis, M.\ 1998, \apj, 500, 525

@@

%\bibitem[Steidel et al.(2010)]{ste10} Steidel, C.~C., Erb, D.~K.,
%  Shapley, A.~E., Pettini, M., Reddy, N., Bogosavljevi{\'c}, M.,
%  Rudie, G.~C., \& Rakic, O.\ 2010, \apj, 717, 289

@@

%\bibitem[Werner et al.(2004)]{wer04} Werner, M.~W., Roellig, T.~L.,
%  Low, F.~J., et al.\ 2004, \apjs, 154, 1
%\bibitem[Wright et al.(2010)]{wri10} Wright, E.~L., Eisenhardt,
%  P.~R.~M., Mainzer, A.~K., et al.\ 2010, \aj, 140, 1868

@@

>>>>

%\documentclass[apj]{emulateapj}
\documentclass[12pt,preprint]{aastex}

@@

\altaffiltext{1}{Center for Astrophysics and Space Sciences,
  University of California, San Diego, La Jolla, CA 92093, USA}
\altaffiltext{2}{Center for Galaxy Evolution Fellow; aleks@ucsd.edu}
\altaffiltext{3}{Department of Astronomy, University of
  Wisconsin-Madison, Madison, WI 53706, USA}
\altaffiltext{4}{Department of Physics and Astronomy, Dartmouth
  College, Hanover, NH 03755, USA}

@@

However, it also plausible that feedback from a compact starburst
could expel gas with such large velocities.  Indeed, there is evidence
for a positive correlation between outflow velocity and starburst
luminosity \citep{mar05,rup05,tre07}, albeit with significant scatter.
Furthermore, \citet{hec11} recently found outflows with maximum
velocities reaching $1500$~\kms\ in a sample of local starbursts with
compact nuclei, and argued that such velocities could be explained by
a wind launched from $r_0\sim100$~pc and driven by feedback from
massive stars and supernovae.

@@

@@

@@

We gathered photometry from the WISE All-Sky Release, the SDSS Seventh
Data Release, and GALEX General Release 6.  We also obtained
$5\times30$~sec dithered exposures at 3.6~$\mu$m and 4.5~$\mu$m for
all sources with the Infrared Array Camera on the Spitzer Space
Telescope as part of GO program 60145.  We used the post--basic
calibrated data to perform aperture photometry on all sources and
point-source photometry on sources in crowded fields.  We show
spectral energy distributions (SEDs) for the three highest $\sigmasfr$
galaxies in Figure~\ref{fig:seds}.

@@

@@

While the SEDs (Figure~\ref{fig:seds}) and optical spectra
(Figure~\ref{fig:spectra}) for our sample indicate that their
bolometric output is dominated by star formation, it is worthwhile to
consider the level of ongoing AGN activity and its effect on our
measurements.  Among the high-$\sigmasfr$ galaxies, the strongest case
for detectable AGN emission can be made for J1506+5402, which has the
most luminous \oiii~$\lambda5007$ line in the HST sample.  This source
also has a weak \nev~$\lambda3426$ emission line, which is associated
with AGN activity \citep{gil10}.  It was observed with the Chandra
X-ray Observatory (proposal ID 11700896) and had four detected counts,
corresponding to a 0.5--8.0~keV X-ray luminosity of
$10^{42.7}$~erg~s$^{-1}$ (see P. Sell et al., in preparation for more
details on the 12/29 galaxies with Chandra observations).  Using the
relationship between 2--10~keV X-ray and 12.3~$\mu$m mid-IR luminosity
for AGNs from \citet{gan09}, one would expect a source with
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{42}$~erg~s$^{-1}$ to
have
$L_{\textnormal{\scriptsize{MIR}}}\approx7\times10^{42}$~erg~s$^{-1}$,
which is 400$\times$ smaller than the observed luminosity for
J1506+5402
($L_{\textnormal{\scriptsize{MIR}}}\approx3\times10^{45}$~erg~s$^{-1}$).
Based on its \oiii\ luminosity ($10^{42.1}$~erg~s$^{-1}$, which may
also include a contribution from star formation) and the calibration
for type 1 AGNs from \citet{hec05}, one would expect an intrinsic
2--10~keV luminosity of
$L_{\textnormal{\scriptsize{X}}}\approx5\times10^{43}$~erg~s$^{-1}$,
suggesting absorption by a factor of $\sim10$ in the X-rays.  However,
even if the X-ray attenuation were a factor of 100, which is the
typical value for local Compton-thick AGNs \citep{dia09}, the expected
mid-IR AGN contribution would only be relevant at the $\lesssim$30\%
level.  We conclude that the bolometric output of the galaxies in our
sample is dominated by star formation and that our results regarding
large $\sigmasfr$ values are not affected by AGN contamination.

@@

We acknowledge useful discussions with and assistance from James Aird,
Brandon Kelly, Dusan Keres, David Law, Alexander Mendez, Kate Rubin,
Art Wolfe, and Stijn Wuyts.  AMD acknowledges support from the
Southern California Center for Galaxy Evolution, a multi-campus
research program funded by the University of California Office of
Research.  Support for HST GO-12272 was provided by NASA through a
grant from STScI.  Support for Spitzer GO-60145 was provided by
contract 1419615 from JPL/Caltech.

@@

@@

@@

@@

@@

@@

@@

@@

@@

@@

\clearpage

\begin{figure}[!t]
\begin{center}
\includegraphics[angle=0,scale=0.7]{f1.ps}
\caption{Left: Observed UV--IR SEDs ($\lrest=0.1$--15~$\mu$m) for
  three galaxies with the largest SFR surface densities
  ($\sigmasfr\approx3000$~\units).  The top two SEDs are offset by 5
  and 10 magnitudes, respectively.  We show stellar population fits to
  the $\lrest=0.1$--3~$\mu$m emission (black solid line) and three
  templates for dust emission \citep[M82 starburst, blue dashed line;
    Arp 220 starburst, green dot-dashed line; obscured quasar, red
    dotted line;][]{pol07} scaled to match the $\lobs=4.6~\mu$m band.
  The starburst templates provide reasonable fits to the $\lobs=12$
  and 22~$\mu$m WISE photometry, while the quasar template does not.
  Right: HST/WFC3 F814W images (probing $\lrest\approx5000$~\AA)
  showing that these galaxies are dominated by a compact nucleus.}
\label{fig:seds}
\end{center}
\end{figure}

\clearpage

\begin{figure}[!t]
\begin{center}
\includegraphics[angle=0,scale=0.7]{f2.ps}
\caption{One-dimensional surface brightness profile for J0905+5759,
  which has the smallest effective radius in the sample.  The observed
  profile is shown as the solid black line, and the profiles of six
  stars in the same image are shown in blue, normalized to the same
  central surface brightness.  The best-fit de Vaucouleurs profile
  with $r_e=100$~pc is shown as a dashed red line, and for comparison
  a broader profile with $r_e=0.04\arcsec=290$~pc is shown as the
  dotted green line.  This galaxy is quite compact, but is more
  extended than a point source.}
\label{fig:profile}
\end{center}
\end{figure}

\clearpage

\begin{figure}[!t]
\begin{center}
\includegraphics[angle=0,scale=0.7]{f3.ps}
\caption{Spectra covering $\lrest=2500$--5200~\AA\ for the three
  galaxies shown in Figure~\ref{fig:seds}.  For clarity the top two
  spectra are offset by $+10$ and $+20$ units.  These spectra are
  dominated by the light of a young stellar population but have
  relatively weak nebular emission lines (\oii~$\lambda3727$,
  H$\beta$~$\lambda4861$, \oiii~$\lambda5007$) and strong
  \mgii~$\lambda\lambda2796,2803$ absorption arising from the
  interstellar medium.  The absense of broad \mgii\ or H$\beta$
  emission lines rules out a significant contribution to the optical
  continuum light from an AGN.  The panels on the right highlight the
  region around the \mgii\ doublet in both wavelength and velocity
  space to illustrate the outflow kinematics.  The spectrum in the
  bottom panel has sufficient spectral resolution
  ($\textnormal{FWHM}\approx8$~\kms) to resolve the intrinsic shape of
  the absorption-line profile, revealing that the gas near the
  centroid velocity ($v=-2470$~\kms) covers the entire galaxy.}
\label{fig:spectra}
\end{center}
\end{figure}

\clearpage

\begin{figure*}[!t]
\begin{center}
\includegraphics[angle=0,scale=.78]{f4.ps} 
\caption{SFR surface densities and stellar masses for the HST--WISE
  sample described in this paper (black circles, symbol size
  proportional to outflow velocity), along with samples of $z<0.3$
  gas-rich mergers (orange squares) and $z\sim1$ star-forming galaxies
  (shown with 68\%, 95\%, and 99.7\% contours; see
  Section~\ref{sec:discussion}).  We mark the empirical threshold for
  launching winds \citep[dotted line,
    $\sigmasfr\approx0.1$~\units;][]{hec02}, the 90th-percentile
  starburst intensity limit from \citet{meu97} (dashed line,
  $\sigmasfr\approx25$~\units), and the Eddington limit from radiation
  pressure on dust grains \citep[solid line,
    $\sigmasfr\approx3000$~\units;][]{mur05,tho05,hop10}.  The
  representative error bar in the top-right portion of the plot
  corresponds to uncertainties of 0.3~dex in $\sigmasfr$ and 0.2~dex
  in stellar mass.  Our HST--WISE sample overlaps with the region
  characterized by gas-rich mergers, and extends to very large SFR
  surface densities near the Eddington limit, suggesting growth that
  is limited by momentum injection from massive stars.}
\label{fig:sigmasfr}
\end{center}
\end{figure*}
